---
title: "Deriving Polygenic Approximation"
author: "Knoblauch Nicholas"
date: 2017-07-12
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->


# Background

##General properties of compound normal

If $$x|\mu \sim N(A\mu,\Sigma)$$ and $$\mu \sim N(\rho,\Lambda)$$ then the marginalized form of $x$ is $$x \sim N(A\rho,A \Lambda A^{T} + \Sigma)$$

## RSS polygenic prior on $\beta$

According to the RSS likelihood:

$$\hat{\beta}|\beta \sim N(SRS^{-1}\beta,SRS)$$ Where $R$ is the population LD matrix, and $S$ is a diagonal matrix with entires $S_{jj}=\frac{1}{\text{se}(\hat{\beta}_j)}$  This means that if $\beta \sim N(0,I\sigma_{\beta})$ , then we can obtain the the marginalized form of $\hat{\beta}$ by substituting $$A=SRS^{-1}$$ $$\rho=0$$ $$\Lambda=I\sigma_\beta$$  and $$\Sigma=SRS$$



$$\hat{\beta} \sim N(0,(SRS^{-1})I\sigma_\beta(SRS^{-1})^{T}+SRS)=N(0,\sigma_\beta SRS^{-2}RS+SRS)$$

